\LARGE \bf Machine Translation for Tibetan Buddhist Texts

### 2021-04-13 paper/main.tex
First draft of discussion

Considering our use of word segmentation for preprocessing our data (as opposed to character segmentation), our results are comparable with, albeit lower than, Tibetan-Chinese models. Lai, Zhao, and Bao ran several different Tibetan-Chinese models on the same data using two different preprocessing methodologies, character segmentation and word segmentation. Their word segmented Transformer model averaged a higher BLEU score than ours, scoring a 38.44\% while our model scored 23.01\%.

>>>>

Considering our use of word segmentation for preprocessing our data (as opposed to character segmentation), our results are comparable with, albeit lower than, Tibetan-Chinese models. Lai, Zhao, and Bao ran several different Tibetan-Chinese models on the same data using two different preprocessing methodologies, character segmentation and word segmentation. Their word segmented Transformer model averaged a higher BLEU score than ours, scoring a 38.44\% while our model scored 23.01\%. While this comparisons may lead to the conclusion that our model is flawed and irredeemable, only the former is correct. There’s no doubt that our model is underperforming in comparison to other transformer models and hardly makes an correct predictions at all. However, this is most likely due to our difference in preprocessing techniques and the linguistic differences and similarities between English, Chinese, and Tibetan.

Unlike English but similar to Chinese, Tibetan words are not clearly separated by spaces, rather, characters which represent words are strung together to make a sentence. This makes preprocessing the data and mapping Tibetan words to their English counterparts a difficult task and it makes preprocessing on characters impossible. We trained our model on a parallel corpus of English and Tibetan sentences instead of segmenting each word since preprocessing these sentences to word-to-word mappings would require an extensive knowledge on the Tibetan language which our team did not have. On top of this, a Chinese-Tibetan model might perform overall better than a Tibetan-English one since Chinese and Tibetan are part of the same language family, and it is likely that characters may map to one another and the same preprocessing techniques can be used on both languages. Therefore, while comparing our model to Lai, Zhao and Bao’s model may be helpful to give a general baseline on how other Tibetan translators are performing, it should not be considered where our model must perform.

Another issue with our data that may be remedied in future iterations is about the corpora we chose to train and test our model on. Since we trained on sentences in our testing set, it is very possible that the model may have read some of the initial characters and assumed that the entire sentence would be a sentence from our training data that began with the same characters. This may explain why our model overpredicts the sentence “Our blessed one” with such a low perplexity. As stated above, this issue might be remedied if we could preprocess our data so that words mapped to words but since we our limited on our ability to do such, we could consider training our model on a different corpus altogether. Instead of using the translated Buddhist Sutras to train our model, we could use a more general Tibetan-English corpus to train our model and then use transfer learning to finetune our model to religious texts. However, at this current time, the translated sutras provided by 84000 are some of the only parallel translations available on the internet.
